Types From Data: \textnormal{Making structured schema-free data formats first-class citizens in F\#}

### 2015-11-17 pldi-2016/paper.tex
Update paper.tex

We start with an informal overview that shows how F\# Data type providers simplify working with 
JSON, XML and CSV. We introduce the necessary aspects of F\# type providers along the 
way. The examples in this section illustrate a number of key properties of the type inference 
algorithm:

@@

\item The inference mechanism prefers records. This is to allow partial typings for extensible
  data formats, and also interacts well with developer tooling --
  most editors provide code completion hints on ``.'' and so types with properties (records)
  are easier to use than types that require pattern matching.

@@

Informally, the safety property (\S\ref{sec:safety}) states that if the inputs
are subtypes of some of the provided samples (i.e.~the samples are representative), then
no exceptions will occur. In other words, we cannot avoid all failures, but we can prevent some.
If the OpenWeatherMap changes the response format, the sample in \S\ref{sec:introduction}
will not re-compile and the user knows that the code needs to be changed. What this means for
type safety is discussed in \S\ref{sec:safety-discuss}.

@@

\paragraph{Open world and annotated \ident{any} types.}
In the above example, we do not infer a union type representing a choice between 
heading, paragraph and image. By its nature, XML is extensible and the sample cannot include all 
possible nodes\footnote{Even when the document structure is defined using XML Schema, documents may 
contain elements prefixed with other namespaces.}. The type of \ident{elem} is thus a top type (we 
call it an \emph{any} type) annotated with the statically known possibilities (heading, paragraph, image). 
It is ultimamtely mapped to the following F\# type:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Element}~=  \\[0.1em]
 \quad \kvd{member}~\ident{Heading}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Paragraph}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Image}~:~\ident{option}\langl \ident{Image} \rangl\\
\end{array}
\end{equation*}
%
This F\# type is worth looking at closely to understand the formalism that follows.  The type
provides statically-typed access to the possible elements known statically from the sample.
However the type is not closed -- the data may be some other element about which nothing is known.
This is the \emph{open-world assumption} -- the user also needs to handle the case when the value is none 
of the statically known elements.  The F\# type simply records the possibility that the value might be one of
the given elements.
%The type is also not exclusive: it doesn't promise that only one of the cases returns a non-\ident{None} value.

As a consequence, each \ident{any} type can be seen as a top type with annotations. The above code will work
correctly on document that contains any elements, including those not listed in the sample. This
is highly useful when reading data from external data sources, because the sample document need only
contain values relevant to the application. The sample need only specify a selective
but sufficient characterization of the open-ended possibilities of extensible data. The type 
inference minimzes the number of variant types that appear in the resulting type --
and a variant is not used if there is an alternative (Corollary~\ref{thm:no-unions}).

@@

\item There is a bottom type (\ref{eq:sub-bot}) and variants behave as top types, because
  any type $\sigma$ is a subtype of any variant (\ref{eq:sub-var-top}). There is a range of
  variants (no \emph{single} top type), but any variant is a subtype of any other variant.

\item As usual, the subtyping on records is covariant (\ref{eq:sub-record1}), subtype can have 
  additional fields (\ref{eq:sub-record2}) and fields can  be reordered (\ref{eq:sub-record3}). 
  The interesting rule is (\ref{eq:sub-record4}); together with covariance, it states that a 
  subtype can omit some fields, provided that their types are nullable.
\end{itemize}

\noindent
If we were type-checking programs, defining variants as a family of top types (which are all 
mutual subtypes) would not add any value over having a single top type, but our purpose is 
different. We simplify reading structured data and, as shown earlier (\S\ref{sec:providers-xml}), 
the types forming a variant provide a valuable information about the sample document.

A particularly important rule is (\ref{eq:sub-record4}), which allows subtype to have fewer record 
elements. This allows us to prefer records over variants. For example, given 
$\{ \ident{name}\!:\!\ident{string} \}$ and $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int} \}$,
we find a common supertype $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int}~\kvd{option} \}$.
This is a least upper bound (clarified below) and more usable than another common supertype 
$\kvd{var}\langl \{ \ident{name}\!:\!\ident{string} \}, \{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int} \}\rangl$.

Some of the aspects of our system (numeric types, handling of \kvd{null} and treatment of
missing data) are based on common F\# practices and might differ for other languages. For example, 
OCaml also has multiple numeric types, but would likely always handle \kvd{null} explicitly.

>>>>

We start with an informal overview that shows how F\# Data type providers simplify working with 
JSON, XML and CSV. We introduce the necessary aspects of F\# type providers along the 
way. The examples in this section illustrate a number of key design principles of the type inference 
algorithm:

@@

\item The inference mechanism prefers types with properties (records). This is to allow extensible (open-world)
  data formats, as we will explain. It also interacts well with developer tooling --
  most editors provide code completion hints on ``.'' and so types with properties (records)
  are easier to use than typessuch as discriminated unions that require pattern matching.

@@

Informally, the safety property (\S\ref{sec:safety}) states that if the inputs
are subtypes of some of the provided samples (i.e.~the samples are representative), then
no exceptions will occur. In other words, we cannot avoid all failures, but we can prevent some.
If OpenWeatherMap changes the response format, the sample in \S\ref{sec:introduction}
will not re-compile and the user knows that the code needs to be changed. What this means for
type safety is discussed in \S\ref{sec:safety-discuss}.

@@

\paragraph{Open world}

In the above example, we do not infer an F\# discriminated union type representing a choice between 
heading, paragraph and image. By its nature, XML is extensible and the sample cannot include all 
possible nodes\footnote{Even when the document structure is defined using XML Schema, documents may 
contain elements prefixed with other namespaces.}. 
The type is not closed -- the data might be some other element about which nothing is known.
This is the \emph{open-world assumption}. Supporting the open-world assumption is crucial when reading data from 
external data sources, because the sample document need only
mention possibilities relevant to the application.

In the subsequent formalism, the type of \ident{elem} will initially be a top type 
annotated with the statically known possibilities (heading, paragraph, image). 
We call this a \emph{variant-annotated top type}.
However, variant-annotated top types are then removed in favour of record types
that have additional optional fields and the type inference minimzes the number of 
variant-annotated top types that appear in the resulting type --
a variant is not used if there is an alternative (Corollary~\ref{thm:no-unions}).

In this case, the variant-annotated top type is ultimamtely presented as the following F\# type:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Element}~=  \\[0.1em]
 \quad \kvd{member}~\ident{Heading}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Paragraph}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Image}~:~\ident{option}\langl \ident{Image} \rangl\\
\end{array}
\end{equation*}
%
This still provides access to the possible elements known statically from the sample.
On the other hand, the user also needs to explicitly handle the casse when the value is none 
of the statically known elements. 
For example, the above code for  \strf{sample.doc} includes a \kvd{None -> ()} case to ensure that it will work 
correctly on a document containing elements not listed in the sample. Similarly, 
the program is not \emph{forced} to consider each of the possiblities in the annotations (in the manner
of complete-coverage checks for discriminated unions). For example,
the program is not forced to cover cases for  \ident{Paragraph} and \ident{Image}.  This is consistent
with our design principles, which prefer records (with optional properties) over discriminated unions.

% Nothing forces the programmer to consider the cases

%The type is also not exclusive: it doesn't promise that only one of the cases returns a non-\ident{None} value.

@@

\item There is a bottom type (\ref{eq:sub-bot}) and variants behave as top types, because
  any type $\sigma$ is a subtype of any variant (\ref{eq:sub-var-top}). There is a range of
  variants (no \emph{single} top type), but any variant is a subtype of variants with fewer tags.

\item As usual, the subtyping on records is covariant (\ref{eq:sub-record1}), subtype can have 
  additional fields (\ref{eq:sub-record2}) and fields can  be reordered (\ref{eq:sub-record3}). 
  Rule (\ref{eq:sub-record4}) is discussed below.
\end{itemize}

\noindent
If we were simply type-checking programs, using variant-annotated top types  
would not add value over having a single top type. However, as shown earlier (\S\ref{sec:providers-xml}), 
the types forming a variant provide a valuable information about the sample document.

A particularly important rule is (\ref{eq:sub-record4}), which, 
together with covariance, states that a supertype can include additional fields, 
provided that their types are nullable (which includes option types). Values of the subtype
will report the null value for these additional fields.
This gives many possible supertypes for a record type: including types with fewer fields and more nullable fields.
This flexiblity will allow us to prefer records-with-extra-nullable-fields
over variants when computing common supertypes. For example, given 
$\{ \ident{name}\!:\!\ident{string} \}$ and $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int} \}$,
we find a common supertype $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int}~\kvd{option} \}$.
This is a least upper bound (clarified below) and preferred over other common supertypes 
$\kvd{var}\langl \{ \ident{name}\!:\!\ident{string} \}, \{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int} \}\rangl$.

Some of the aspects of our system such as the relationship between numeric types is
based on common industry practice and are to some extent tunable. For example, 
options can be added to always assume that all fields are always nullable.  The default
options may differ according to common practice in different programming communities.
